# Employment

https://www.bls.gov/sae/

Benchmark https://www.bls.gov/news.release/presae.nr0.htm
https://www.bls.gov/news.release/presae.tn.htm
https://www.bls.gov/web/empsit/cesbmart.htm national
https://www.bls.gov/sae/publications/benchmark-article/annual-benchmark-article.htm state and area

The State Employment and Unemployment data for January 2026 are expected to be released in April 2026.
https://www.bls.gov/sae/


Phillippa Dunne

## Setup

```{r}
#| label: setup

rm(list = ls())
GITROOT <- rprojroot::find_root(rprojroot::has_file(".git/index"))
source(fs::path(GITROOT, "setup.R"))

```


## Get BLS employment data via BLSloadR and save streamlined version

```{r}
#| eval: false

# ONETIME

# https://cran.r-project.org/web/packages/BLSloadR/BLSloadR.pdf
# library("BLSloadR")

ces_data <- BLSloadR::get_ces() # takes a while
fname <- paste0("ces_data_", Sys.Date(), ".rds")
saveRDS(ces_data, fs::path(DCES, fname))

```

```{r}
#| eval: false

# ONETIME

# time consuming
fname <- "ces_data_2026-02-03.rds"
system.time(ces <- readRDS(fs::path(DCES, fname))) # 19 secs

```


```{r}
#| label: create-compact tibble
#| eval: false

# ONETIME get rid of unnecessary stuff

str(ces) # data.table, data.frame
names(attributes(ces))
ces2 <- attr(ces, "bls_diagnostics")$data |> as_tibble()
attributes(ces2)
saveRDS(ces2, fs::path(DCES, "ces_tibble.rds"))

# get some stats
count(ces2, seasonal)
#    seasonal       n
#      <char>   <int>
# 1:        S 1162521
# 2:        U 8287787

count(ces2, state_code, state_name)
tmp <- count(ces2, industry_code, industry_name) # 00000000 is total nonfarm
tmp2 <- count(ces2, data_type_code, data_type_text) # 01 is all employees

```

### Get compact data

```{r}

system.time(ces <- readRDS(fs::path(DCES, "ces_tibble.rds"))) # somewhat faster 8.7  secs

```

### Save CES states subset for faster loading

```{r}
#| label: ces-states-subset

# create a states total employment file

fips <- tidycensus::fips_codes |>
  distinct(state_code, state) |> # state_code is two-digit string, state is postal abbr
  rename(stabbr = state) |>
  as_tibble()

ces_emp_states <- ces |>
  filter(
    area_code == "00000",
    industry_code == "00000000",
    data_type_code == "01"
  ) |>
  left_join(fips, by = join_by(state_code)) |>
  select(series_id, stabbr, state_code, state_name, seasonal, date, value)
count(ces_emp_states, stabbr, state_code, state_name)

saveRDS(ces_emp_states, fs::path(DCES, "ces_emp_states.rds"))

```


```{r}
#| label: look-states

system.time(ces_states <- readRDS(fs::path(DCES, "ces_emp_states.rds")))

update_geom_defaults("line", list(linewidth = 1))
update_geom_defaults("point", list(size = 1))
ces_emp_states |>
  filter(stabbr == "NY") |>
  filter(year(date) >= 2022) |>
  ggplot(aes(date, value, colour = seasonal)) +
  geom_line() +
  geom_point()

ces_emp_states |>
  filter(stabbr %in% c("NY", "NJ", "MA", "CT", "PA"), seasonal == "S") |>
  filter(year(date) >= 2023) |>
  mutate(
    ivalue = value / value[date == min(date)] * 100,
    .by = c(stabbr, seasonal)
  ) |>
  ggplot(aes(date, ivalue, colour = stabbr)) +
  geom_line() +
  geom_point()

update_geom_defaults("line", list(linewidth = 2))
update_geom_defaults("point", list(size = 3))

```

### Save CES NY subset for faster loading

```{r}
#| label: ces-ny-subset

inds <- count(ces, supersector_code, industry_code, industry_name)
count(ces, data_type_code, data_type_text)

cesny <- ces |>
  filter(
    state_code == "36",
    data_type_code == "01"
  ) |>
  select(-c(state_code, state_name, data_type_code, data_type_text)) |>
  relocate(value, .after = date)

saveRDS(cesny, fs::path(DCES, "cesny.rds"))

```


```{r}
#| label: cesny-look
#| eval: false

glimpse(cesny)

cesny |>
  filter(year(date) >= 2023) |>
  filter(area_code == "00000", supersector_code == "00") |>
  ggplot(aes(date, value, colour = seasonal)) +
  geom_line() +
  geom_point()


```


## New York Fed early benchmark data

```{r}
#| label: nyfed-early-benchmark

empnyfed1 <- vroom::vroom(fs::path(
  DFEDNY,
  "early-benchmarked-employment_data.csv"
))

empnyfed2 <- empnyfed1 |>
  rename_with(tolower) |>
  pivot_longer(
    cols = -date,
    names_to = "area",
    values_to = "emp"
  ) |>
  mutate(
    date = lubridate::mdy(date),
    date = floor_date(date, "month"),
    type = "early_bench",
    area = str_remove(area, "_bench")
  ) |>
  select(date, type, area, emp)

saveRDS(empnyfed2, fs::path(DFEDNY, "early-benchmarked-employment_data.rds"))

empnyfed2 |>
  filter(area == "ny") |>
  ggplot(aes(date, emp)) +
  geom_line() +
  geom_point() +
  theme_presentation()

```

## CES data from NY State data.ny.gov -- NOT implemented

```{r}
#| label: ces-data.ny.gov
#| eval: false

# https://data.ny.gov/Economic-Development/Current-Employment-Statistics-Beginning-1990/6k74-dgkb
# url <- "https://data.ny.gov/a1177f73-6628-4654-b8d3-25cc2335bf5b"
url <- "https://data.ny.gov/api/views/6k74-dgkb/rows.csv?accessType=DOWNLOAD" # file size is smaller than manual download because it has numeric not characer data
fname <- "Current_Employment_Statistics__Beginning_1990.csv"

# fname <- "cesny_data.csv"
download.file(url, destfile = fs::path(DCESNY, fname))

```

```{r}
#| eval: false

# These data are NS except for NYS totals
fname <- "Current_Employment_Statistics__Beginning_1990.csv"
cnames <- c("area", "areaname", "series", "title", "year", "month", "emp")
cesny <- vroom::vroom(fs::path(DCESNY, fname))

cesny1 <- cesny |>
  set_names(cnames) |>
  mutate(
    rectype = ifelse(month == 0, "annual", "month"),
    date = make_date(year, month, 1)
  ) |>
  select(-month)

glimpse(cesny1)
skimr::skim_without_charts(cesny1)

count(cesny1, area, areaname) # nys 00000; nyc 93561; 28 counties; 13 areas
# count(cesny1, series, title) # 191 series codes in total -- main level:
#    series   title                                                                              n
#    <chr>    <chr>                                                                          <int>
#  1 00000000 Total Nonfarm                                                                  19518
#  2 05000000 Total Private                                                                  19518
#  3 06000000 Goods Producing                                                                19518
#  4 07000000 Service-Providing                                                              19518
#  5 08000000 Private Service Providing                                                      19518

```

```{r}
#| eval: false

cesny1 |>
  filter(area == "00000", series == "00000000", rectype == "annual") |>
  ggplot(aes(year, emp)) +
  geom_line() +
  geom_point() +
  theme_minimal()

cesny1 |>
  filter(
    area == "00000",
    series == "00000000",
    rectype == "month",
    year >= 2022
  ) |>
  ggplot(aes(date, emp)) +
  geom_line() +
  geom_point() +
  theme_minimal()

cesny1 |>
  filter(area == "00000", series == "00000000", rectype == "month") |>
  arrange(date) |>
  mutate(pchya = emp / lag(emp, 12) - 1, .by = c(area, series, rectype)) |>
  filter(year >= 2023) |>
  ggplot(aes(date, pchya)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "solid", color = "blue") +
  theme_minimal() +
  scale_y_continuous(
    breaks = seq(-1, 1, .005),
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  scale_x_date(
    name = NULL,
    date_labels = "%b-%Y",
    breaks = seq(as.Date("2023-01-01"), as.Date("2025-12-01"), by = "3 months")
  ) +
  labs(
    title = "New York State Employment Year-over-Year Percent Change",
    y = "% change"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

cesny1 |>
  filter(area == "00000", series == "00000000", rectype == "month") |>
  arrange(date) |>
  mutate(pchya = emp / lag(emp, 12) - 1, .by = c(area, series, rectype)) |>
  filter(year >= 2023)

```


## LAUS data via BLSloadR -- not yet implemented

```{r}
#| eval: false

laus_states <- get_laus()
laus_states_unadjusted <- get_laus("state_unadjusted")

laus_states_unadjusted |>
  filter(
    area_text == "New York",
    measure_text == "employment",
    year(date) >= 2022
  )

tmp <- laus_states |>
  filter(area_text == "New York")

tmp |>
  filter(measure_text == "employment") |>
  ggplot(aes(x = date, y = value)) +
  geom_line() +
  geom_point() +
  theme_minimal()

```
