
# Get QCEW data

-   2023 is the latest full year as of 3/22/325
-   Get annual QCEW data for NY counties
-   Do minor cleaning
-   Put NYC and MTA markers on the file
-   Save as `qcewstco.rds` in the intermediate data folder
-   Save an MTA subset with just the latest year

## QCEW data from the New York State Department of Labor

-   [NYSDOL QCEW landing page](https://dol.ny.gov/quarterly-census-employment-and-wages)
-   [direct zip link](https://statistics.labor.ny.gov/qcew.zip)
-   [technical notes](https://dol.ny.gov/qcew-technical-notes)

The New York State Department of Labor (NYSDOL) provides a version of the BLS Quarterly Census of Employment and Wages (QCEW) that has quarterly county-level wage, establishment, and employment data in a format we need. The data are available from 2000 through a recent quarter. A new quarter and revisions to prior quarters generally are available about 6 months after end of a quarter.

## Setup

```{r}
#| label: setup

rm(list = ls())
GITROOT <- rprojroot::find_root(rprojroot::has_file(".git/index"))
source(fs::path(GITROOT, "setup.R"))

```



```{r}
#| label: qcew-names
#| output: false

url <- "https://statistics.labor.ny.gov/qcew.zip"
fname <- fs::path_file(url)
fname <- "qcew_2026-02-02.zip"

zpath <- fs::path(DQCEW, fname)
# zpath <- paste0(DQCEW, "/", fname)

```

## Download qcew data

Note: The latest qcew.zip file, which I downloaded once manually and once with the code below, appears to have been corrupted, at least as downloaded by my machine. I had to use an external utility (WinRar) to unzip it to a separate folder, then rezip it, after which the computer could read it.

Change eval: to true in chunk below to re-download data.

```{r}
#| label: download-qcew
#| eval: false

# Sys.setenv(R_DEFAULT_INTERNET_TIMEOUT = 400)
# options(timeout = max(400, getOption("timeout")))
# download.file(url, zpath, mode = "wb")

```

```{r}
#| label: make-factors
#| output: false

# I don't know what codes NYSDOL uses so make my own
# ownf <- c(`1`="Total Private and Government",
#           `2`="Private",
#           `3`="Total Government",
#           `4`="Federal Government",
#           `5`="State Government",
#           `6`="Local Government")

owner_factors <- vroom(
  I(
    "
ownerf, owner, ownershort
1, Total Private and Government, totown
2, Private, privown
3, Total Government, totgovown
4, Federal Government, fedgovown
5, State Government, stategovown
6, Local Government, localgovown
7, Other Government, othergovown
"
  )
)
write_csv(owner_factors, fs::path(DDATA, "owner_factors.csv"))

```


## Quarterly data

```{r}
#| label: get-save-qcew-state-county
#| eval: true

zfiles <- zip::zip_list(zpath)

# csvfiles <- zfiles |>
#   filter(str_starts(filename, "qcew_annual_")) |>
#   pull(filename) |>
#   sort()

csvfiles <- zfiles |>
  filter(str_starts(filename, "qcew_quarter_")) |>
  pull(filename) |>
  sort()

system.time(df <- vroom(purrr::map(csvfiles, \(x) unz(zpath, x)))) # 10.3 secs; unzipping all and using purrr is faster

df2 <- df |>
  rename_with(str_to_lower)

# count(df2, areatype)
# count(df2, areatype, owner)

df3 <- df2 |>
  filter(areatype %in% c("State", "County")) |>
  left_join(owner_factors, by = join_by(owner)) |>
  mutate(
    areatype = str_to_lower(areatype),
    empq = pmean(mnth1emp, mnth2emp, mnth2emp),
    date = yq(paste(year, quarter, sep = "-"))
  )
glimpse(df3)
count(df3, ownerf, ownershort, owner)
summary(df3)

df4 <- df3 |>
  select(
    date,
    areatype,
    area,
    naics,
    naics_level,
    naics_title,
    ownerf,
    owner,
    ownershort,
    mnth1emp,
    mnth2emp,
    mnth2emp,
    empq,
    totwage
  )
summary(df4)
# skimr::skim_without_charts(df4)

saveRDS(df4, fs::path(DQCEW, "qcewstcoq.rds"))

```


```{r}

qcew <- readRDS(fs::path(DQCEW, "qcewstcoq.rds"))
summary(qcew)
glimpse(qcew)
count(qcew, ownerf, ownershort)

tmp <- qcew |>
  filter(ownershort == "totown", areatype == "state")

tmp |>
  filter(naics == "00") |>
  arrange(date) |>
  mutate(pchya = totwage / lag(totwage, 4) - 1) |>
  ggplot(aes(date, pchya)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(
    breaks = seq(-.5, .5, .01),
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_bw()

tmp |>
  filter(naics == "00") |>
  arrange(date) |>
  mutate(pchya = empq / lag(empq, 4) - 1) |>
  ggplot(aes(date, pchya)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(
    breaks = seq(-.5, .5, .01),
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_bw()


```


```{r stop_here, echo=FALSE}
knitr::knit_exit()
```