---
title: "New York Tax Revenue and the Economy"
subtitle: "NYS Consensus Economic and Revenue Forecasting Conference"
author: "Don Boyd"
date: "2026-02-23"
date-format: long
format:
  revealjs:
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    # logo: images/quarto.png
    # css: styles.css
    # footer: '[https://{{< meta prerelease-subdomain >}}quarto.org](https://{{< meta prerelease-subdomain >}}quarto.org)'
  # pdf:
  #   output-file: slides.pdf
number-sections: false
# resources:
#   - demo.pdf
# quarto render index.qmd --to pdf
# quarto render index.qmd --to revealjs
---

## Key revenue forecasting questions

-   What will they owe?
-   When will they pay it?

Often it's important to consider these questions separately.


```{r}
#| label: setup

rm(list = ls())
GITROOT <- rprojroot::find_root(rprojroot::has_file(".git/index"))
DDATA <- fs::path(GITROOT, "data")
DQCEW <- fs::path("D:/", "data", "qcew")

source(fs::path(GITROOT, "libraries.R"))
source(fs::path(GITROOT, "libraries_ts.R"))
source(fs::path(GITROOT, "plot_helpers.R"))

# QROOT <- fs::path(GITROOT, "tmd", "national_targets", "qmd")
# source(fs::path(QROOT, "setup.R"))

```

```{r}

pop <- readRDS(fs::path(DDATA, "ny_state_population_estimates_1970_2024.rds"))

```


## Major challenges

-  A cloudy rear-view mirror -- hard to know the future when we don't know the present
-  Forecasters stand on shifting sands -- economic data sometimes are revised substantially
-  The economy we measure is not (quite) the economy we tax

# Where we are now

## Population

```{r}

p1 <- pop |>
  filter(variable == "popestimate") |>
  ggplot(aes(year, value)) +
  geom_line(colour = "blue") +
  scale_y_continuous(
    name = "millions",
    labels = scales::label_comma(scale = 1e-6, accuracy = .1)
  ) +
  scale_x_continuous(name = NULL) +
  ggtitle("New York State Post-Census Population Estimates") +
  labs(caption = "Source: U.S. Census Bureau Population Estimates Program") +
  theme_presentation_small()

p2 <- p1
p3 <- p1
p4 <- p1

```


```{r}

f <- function(component) {
  p <- pop |>
    filter(variable == component) |>
    ggplot(aes(year, value)) +
    geom_line(colour = "blue") +
    scale_y_continuous(
      name = "millions",
      labels = scales::label_comma(scale = 1e-6, accuracy = .1)
    ) +
    scale_x_continuous(name = NULL) +
    ggtitle("New York State Post-Census Population Estimates") +
    labs(caption = "Source: U.S. Census Bureau Population Estimates Program") +
    theme_presentation_small()
  p
}

f("naturalchg")

```


```{r}
#| eval: false

# keep <- c(
#   "popestimate",
#   "naturalchg",
#   "domesticmig",
#   "internationalmig",
#   "npopchg",
#   "residual"
# )

keep <- c(
  "naturalchg",
  "domesticmig",
  "internationalmig"
)

pop |>
  filter(variable %in% keep) |>
  ggplot(aes(x = factor(year), y = value, fill = variable)) +
  geom_col(position = "dodge") +
  labs(
    title = "New York State Population Change Components",
    x = "Year",
    y = "Population Change",
    fill = "Component"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

pop |>
  filter(variable %in% keep) |>
  ggplot(aes(x = variable, y = value, fill = factor(year))) +
  geom_col(position = "dodge") +
  labs(
    title = "New York State Population Change Components",
    x = "Year",
    y = "Population Change",
    fill = "Component"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r}
#| label: four-plots
#| fig-width: 18
#| fig-height: 10

(p1 + p2) / (p3 + p4)
```

## Employment

## Revenue

# Another section

## The economy we tax

Table or bar chart -- AGI and personal income

## Risks

- how big? / how wrong?
- when will we know

## Why capital gains are important

- large
- concentrated
- hot money - discretionary
- payment timing

## Large

## Concentrated

## Hot money

## Timing
