---
title: "New York Tax Revenue and the Economy"
subtitle: "NYS Consensus Economic and Revenue Forecasting Conference"
author: "Don Boyd"
date: "2026-02-23"
date-format: long
format:
  revealjs:
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    # logo: images/quarto.png
    # css: styles.css
    # footer: '[https://{{< meta prerelease-subdomain >}}quarto.org](https://{{< meta prerelease-subdomain >}}quarto.org)'
  # pdf:
  #   output-file: slides.pdf
number-sections: false
bibliography: references.bib
nocite: '@*'
execute:
  output: false
# quarto render index.qmd --to pdf
# quarto render index.qmd --to revealjs
---

## Key revenue forecasting questions

-   What will they owe?
-   When will they pay it?

Often it's important to consider these questions separately.


```{r}
#| label: setup

rm(list = ls())
GITROOT <- rprojroot::find_root(rprojroot::has_file(".git/index"))
source(fs::path(GITROOT, "setup.R"))

```

```{r}
#| label: get-data

pop <- readRDS(fs::path(DDATA, "ny_state_population_estimates_1970_2024.rds"))
nyspi <- readRDS(fs::path(DDATA, "bea_nyspi.rds"))
nysagi <- readRDS(fs::path(DPITNY, "nypitds2.rds"))
qcew <- readRDS(fs::path(DQCEW, "qcewstcoq.rds")) # takes a long time, winnow it down
ces <- readRDS(fs::path(DCES, "ces_data_2026-02-03.rds")) # same

```


## Major challenges

-  A cloudy rear-view mirror -- hard to know the future when we don't know the present
-  Forecasters stand on shifting sands -- economic data sometimes are revised substantially
-  The economy we measure is not (quite) the economy we tax

# Where we are now


## Post-Census Population Change

```{r}
#| label: pop-change-facets
#| output: true
#| fig-width: 18
#| fig-height: 10

keep <- c(
  "npopchg",
  "naturalchg",
  "domesticmig",
  "internationalmig"
)
labels <- c(
  "Net Population Change",
  "Natural Change (Births minus Deaths)",
  "Domestic Migration",
  "International Migration"
)

pop |>
  filter(variable %in% keep) |>
  filter(year > 2020) |>
  mutate(variable = factor(variable, levels = keep, labels = labels)) |>
  arrange(variable) |>
  ggplot(aes(x = factor(year), y = value, fill = variable)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(
      "Net Population Change" = "steelblue",
      "Natural Change (Births minus Deaths)" = "seagreen",
      "Domestic Migration" = "coral",
      "International Migration" = "mediumpurple"
    )
  ) +
  geom_hline(yintercept = 0) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(
    name = "change (thousands)",
    labels = scales::label_comma(scale = 1e-3, accuracy = 1),
    limits = c(-350e3, 350e3),
    breaks = seq(-400e3, 400e3, 100e3)
  ) +
  ggtitle("Annual Changes in New York State Population Since 2020 Census") +
  labs(caption = "Source: U.S. Census Bureau Population Estimates Program") +
  facet_wrap(~variable, ncol = 2, scales = "fixed") +
  theme_presentation() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = rel(1.3), face = "bold")
  )

```


# Linking economic and tax-revenue forecasts

```{r}
#| label: get-agi-pi-data

# nysagi
# nyspi

agi2 <- nysagi |>
  filter(
    rstat == "All Filers",
    por == "All Places",
    county == "All Places",
    state == "All",
    country == "All"
  ) |>
  select(year, sortincome, agirange, numret:nyidadjust) |>
  arrange(sortincome, year)

agibase <- agi2 |>
  filter(sortincome == 16) |>
  select(year, nysagi) |>
  mutate(variable = "nysagi", value = nysagi) |>
  select(variable, year, value)

# agi2 |> filter(sortincome == 16) |> ggplot(aes(year, nysagi)) + geom_line()

pibase <- nyspi |>
  filter(linecode == 10, year %in% 2015:2023) |>
  mutate(
    variable = "nyspi",
    year = as.integer(year),
    value = as.numeric(value) * 1e6
  ) |>
  select(variable, year, value)

piagi <- bind_rows(pibase, agibase)

# piagi |> ggplot(aes(year, value, colour = variable)) + geom_line()
# piagi |>
#   pivot_wider(names_from = variable, values_from = value) |>
#   mutate(ratio = nyspi / nysagi) |>
#   ggplot(aes(year, ratio)) +
#   geom_line()
# pi is about 30-50% larger than agi depending on year

pdata <- piagi |>
  arrange(variable, year) |>
  mutate(pchya = value / lag(value) - 1, .by = variable) |>
  filter(year %in% 2016:2023)

```

<!-- ## Economic and tax-base variables often diverge -->

## <span style="font-size: 0.8em;">Economic and tax-base variables often diverge</span>

```{r}
#| label: piagi-growth-plot
#| output: false
#| fig-width: 18
#| fig-height: 10

pdata |>
  mutate(variable = factor(variable, levels = c("nyspi", "nysagi"))) |>
  ggplot(aes(year, pchya, colour = variable)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_color_manual(
    values = c("nysagi" = "#0072B2", "nyspi" = "#009E73"), #  "#E69F00"
    labels = c("nysagi" = "Adjusted Gross Income", "nyspi" = "Personal Income")
  ) +
  scale_y_continuous(
    name = "% change",
    breaks = seq(-.1, .2, .02),
    labels = scales::label_percent(accuracy = 0.1)
  ) +
  scale_x_continuous(name = NULL, breaks = 2015:2024) +
  ggtitle("New York State Personal Income and AGI Growth Rates") +
  labs(caption = "Sources: BEA and NYS Department of Taxation and Finance") +
  theme_presentation() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.85, 0.85),
    legend.background = element_rect(fill = "white", color = "gray80"),
    legend.key.size = unit(1.5, "lines")
  )

```


## AGI vs. Personal Income Components

```{r}
#| label: get-agi-components

# nysagi
# nyspi
# wages, interest, dividends, capgains, caploss, rrpetgain, rrpetloss, busfarmgain, bus

agi_components <- nysagi |>
  filter(
    rstat == "All Filers",
    por == "All Places",
    county == "All Places",
    state == "All",
    country == "All"
  ) |>
  filter(sortincome == 16) |>
  select(
    year,
    numret,
    nysagi,
    wages,
    interest,
    dividends,
    capgains,
    caploss,
    rrpetgain,
    rrpetloss,
    busfarmgain,
    busfarmloss,
    pensions,
    otherinc
  ) |>
  arrange(year)

agibase <- agi_components |>
  mutate(
    intdiv = interest + dividends,
    capnet = capgains - caploss,
    rrpetnet = rrpetgain - rrpetloss,
    busfarmnet = busfarmgain - busfarmloss,
    allother = nysagi -
      wages -
      intdiv -
      capnet -
      rrpetnet -
      busfarmnet -
      pensions -
      otherinc
  ) |>
  select(
    year,
    nysagi,
    wages,
    intdiv,
    capnet,
    rrpetnet,
    busfarmnet,
    pensions,
    otherinc,
    allother
  ) |>
  arrange(year)

agibase2 <- agibase |>
  mutate(other = nysagi - wages - capnet) |>
  select(year, nysagi, wages, capnet, other)

# combine agi components plus nyspi
piagi <- agibase2 |>
  pivot_longer(cols = -year, names_to = "variable", values_to = "value") |>
  bind_rows(pibase)

# vars <- c("nyspi", "nysagi", "wages", "capnet", "other")
# piagi |>
#   filter(variable != "other") |>
#   arrange(variable, year) |>
#   mutate(
#     pchya = value / lag(value) - 1,
#     .by = variable,
#     variable = factor(variable, levels = vars)
#   ) |>
#   arrange(variable, year) |>
#   ggplot(aes(year, pchya, colour = variable)) +
#   geom_line() +
#   geom_point() +
# geom_hline(yintercept = 0)

# agibase |> mutate(check=wages+intdiv+capnet+rrpetnet+busfarmnet+pensions+otherinc+allother - nysagi) |> summarize(max_abs_check = max(abs(check)))

# agibase2 |>
#   mutate(across(-year, \(x) x - lag(x)))

# agibase2 |>
#   mutate(across(-year, \(x) x - lag(x))) |>
#   mutate(across(-c(year, nysagi), \(x) x / nysagi))

```


```{r}
#| label: show-pi-agi-components
#| output: false
#| eval: false
#| fig-width: 18
#| fig-height: 10

vars <- c("nyspi", "nysagi", "wages", "capnet")
var_labels <- c(
  "nyspi" = "Personal Income",
  "nysagi" = "Adjusted Gross Income",
  "wages" = "Wages & Salaries",
  "capnet" = "Net Capital Gains"
)
var_colors <- c(
  "nyspi" = "#009E73", # green
  "nysagi" = "#0072B2", # blue
  "wages" = "#E69F00", # orange
  "capnet" = "#D55E00" # red
)

piagi |>
  filter(variable != "other") |>
  arrange(variable, year) |>
  mutate(
    pchya = value / lag(value) - 1,
    .by = variable,
    variable = factor(variable, levels = vars)
  ) |>
  filter(!is.na(pchya)) |>
  ggplot(aes(year, pchya, colour = variable)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_color_manual(
    values = var_colors,
    labels = var_labels
  ) +
  scale_y_continuous(
    name = "% change",
    breaks = seq(-.6, .6, .1),
    labels = scales::label_percent(accuracy = 1)
  ) +
  scale_x_continuous(name = NULL, breaks = 2015:2024) +
  ggtitle("Growth Rates of Personal Income, AGI, and Major Components") +
  labs(caption = "Sources: BEA and NYS Department of Taxation and Finance") +
  theme_presentation() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.85, 0.85), # upper left to avoid data
    legend.background = element_rect(fill = "white", color = "gray80"),
    legend.key.size = unit(1.5, "lines")
  )
```


```{r}
#| label: show-pi-agi-components2
#| output: false
#| eval: false
#| fig-width: 18
#| fig-height: 10

vars <- c("nyspi", "nysagi", "wages", "capnet", "other")
var_labels <- c(
  "nyspi" = "Personal Income",
  "nysagi" = "Adjusted Gross Income",
  "wages" = "Wages & Salaries",
  "capnet" = "Net Capital Gains",
  "other" = "Other AGI Components"
)
var_colors <- c(
  "nyspi" = "#009E73", # green
  "nysagi" = "#0072B2", # blue
  "wages" = "#E69F00", # orange
  "capnet" = "#D55E00", # red
  "other" = "gray50"
)

piagi |>
  # filter(variable != "other") |>
  arrange(variable, year) |>
  mutate(
    pchya = value / lag(value) - 1,
    .by = variable,
    variable = factor(variable, levels = vars)
  ) |>
  filter(!is.na(pchya)) |>
  ggplot(aes(year, pchya, colour = variable)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_color_manual(
    values = var_colors,
    labels = var_labels
  ) +
  scale_y_continuous(
    name = "% change",
    breaks = seq(-.6, .6, .1),
    labels = scales::label_percent(accuracy = 1)
  ) +
  scale_x_continuous(name = NULL, breaks = 2015:2024) +
  ggtitle("Growth Rates of Personal Income, AGI, and Major Components") +
  labs(caption = "Sources: BEA and NYS Department of Taxation and Finance") +
  theme_presentation() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.85, 0.85), # upper left to avoid data
    legend.background = element_rect(fill = "white", color = "gray80"),
    legend.key.size = unit(1.5, "lines")
  )

```


```{r}
#| label: show-pi-agi-components3
#| output: true
#| fig-width: 18
#| fig-height: 10

vars <- c("nyspi", "nysagi", "wages", "capnet")
var_labels <- c(
  "nyspi" = "Personal Income",
  "nysagi" = "Adjusted Gross Income",
  "wages" = "Wages & Salaries",
  "capnet" = "Net Capital Gains"
)
var_colors <- c(
  "nyspi" = "#009E73", # green
  "nysagi" = "#0072B2", # blue
  "wages" = "#E69F00", # orange
  "capnet" = "#D55E00" # red
)
var_linewidths <- c(
  "nyspi" = 2,
  "nysagi" = 2,
  "wages" = 1,
  "capnet" = 1
)

piagi |>
  filter(variable != "other") |>
  arrange(variable, year) |>
  mutate(
    pchya = value / lag(value) - 1,
    .by = variable,
    variable = factor(variable, levels = vars)
  ) |>
  filter(!is.na(pchya)) |>
  ggplot(aes(year, pchya, colour = variable)) +
  geom_line(aes(linewidth = variable)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_color_manual(
    values = var_colors,
    labels = var_labels
  ) +
  scale_linewidth_manual(
    values = var_linewidths,
    guide = "none" # hide linewidth from legend
  ) +
  scale_y_continuous(
    name = "% change",
    breaks = seq(-.6, .6, .1),
    labels = scales::label_percent(accuracy = 1)
  ) +
  scale_x_continuous(name = NULL, breaks = 2015:2024) +
  ggtitle("Growth Rates of Personal Income, AGI, and Major Components") +
  labs(caption = "Sources: BEA and NYS Department of Taxation and Finance") +
  theme_presentation() +
  theme(
    legend.title = element_blank(),
    legend.position = c(0.85, 0.85),
    legend.background = element_rect(fill = "white", color = "gray80"),
    legend.key.size = unit(1.5, "lines")
  )

```


## Employment

```{r}

qcew
ces

qcew2 <- qcew |>
  filter(areatype == "state", naics == "00", ownerf == 1) |>
  select(qdate = date, starts_with("mnth")) |>
  pivot_longer(cols = -qdate) |>
  mutate(monthadd = str_sub(name, 5, 5) |> as.integer() - 1) |>
  mutate(date = qdate + months(monthadd), src = "qcew") |>
  select(date, src, value)

ces2 <- ces |>
  filter(
    state_code == "36",
    area_code == "00000",
    supersector_code == "00",
    data_type_code == "01"
  ) |>
  mutate(src = ifelse(seasonal == "U", "ces_ns", "ces_s")) |>
  select(date, src, value)

skimr::skim_without_charts(ces2)

qcew2 |>
  ggplot(aes(date, value)) +
  geom_line() +
  geom_point()

comp <- bind_rows(qcew2, ces2) |>
  filter(year(date) >= 2000)

comp |>
  ggplot(aes(date, value, colour = src)) +
  geom_line() +
  geom_point()

wide <- comp |>
  pivot_wider(names_from = src, values_from = value) |>
  mutate(diff = ces_ns - qcew, pdiff = diff / qcew)

wide |>
  ggplot(aes(date, diff)) +
  geom_line() +
  geom_point()

wide |>
  ggplot(aes(date, pdiff)) +
  geom_line() +
  geom_point()

library(feasts)
library(tsibble)

# Convert to tsibble (time series tibble)
# Convert to tsibble with monthly interval
wide_ts <- wide |>
  filter(!is.na(pdiff)) |>
  mutate(month = yearmonth(date)) |>
  as_tsibble(index = month)

# Check for gaps (should be none now)
scan_gaps(wide_ts)

scan_gaps(wide_ts)

wide_ts |>
  filter(year(date) >= 2021) |>
  ggtime::gg_season(pdiff, period = "year") +
  labs(
    title = "Seasonal Pattern of QCEW vs CES Employment Difference",
    y = "Percentage Difference",
    x = "Month"
  )

comp |>
  filter(year(date) >= 2022, src == "ces_ns") |>
  mutate(month = yearmonth(date)) |>
  as_tsibble(index = month) |>
  ggtime::gg_season(value, period = "year") +
  labs(
    title = "Seasonal Pattern of CES Employment",
    y = "Employment",
    x = "Month"
  )

comp |>
  filter(year(date) >= 2022, src != "qcew") |>
  ggplot(aes(date, value, colour = src)) +
  geom_line() +
  geom_point()

comp |>
  arrange(date) |>
  mutate(pchya = value / lag(value, 12) - 1, .by = src) |>
  # filter(year(date) >= 2022, src != "qcew") |>
  filter(year(date) >= 2022) |>
  ggplot(aes(date, pchya, colour = src)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(
    name = "% change",
    labels = scales::label_percent(accuracy = 0.1),
    breaks = seq(-.1, .2, .005),
    limits = c(0, NA)
  ) +
  ggtitle("NYS employment % change vs. year ago") +
  scale_x_date(name = NULL, date_breaks = "3 months", date_labels = "%b %Y") +
  theme_presentation() +
  theme_rotate(angle = 45)

comp |>
  filter(year(date) %in% 2024:2025, month(date) %in% 1:6) |>
  mutate(year = year(date)) |>
  summarise(value = mean(value), .by = c(year, src)) |>
  pivot_wider(names_from = year, names_prefix = "y", values_from = value) |>
  mutate(pch = (y2025 - y2024) / y2024)

bind_rows(qcew2, ces2) |>
  filter(year(date) >= 2022) |>
  ggplot(aes(date, value, colour = src)) +
  geom_line() +
  geom_point() +
  scale_x_date(name = NULL, date_breaks = "3 months", date_labels = "%b %Y") +
  scale_y_continuous(
    name = "Employment (millions)",
    labels = scales::label_comma(scale = 1e-6, accuracy = 0.1),
    breaks = seq(7e6, 10e6, 1e5)
  ) +
  theme_presentation() +
  theme_rotate(angle = 45)

```

## Revenue

# Another section

## The economy we tax

Table or bar chart -- AGI and personal income

## Risks

- how big? / how wrong?
- when will we know

## Why capital gains are important

- large
- concentrated
- hot money - discretionary
- payment timing

## Large

## Concentrated

## Hot money

## Timing

## References

::: {#refs style="font-size: 0.7em;"}
:::
