---
title: "New York Tax Revenue and the Economy"
subtitle: "NYS Consensus Economic and Revenue Forecasting Conference"
author: "Don Boyd"
date: "2026-02-23"
date-format: long
format:
  revealjs:
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    # logo: images/quarto.png
    # css: styles.css
    # footer: '[https://{{< meta prerelease-subdomain >}}quarto.org](https://{{< meta prerelease-subdomain >}}quarto.org)'
  # pdf:
  #   output-file: slides.pdf
number-sections: false
# resources:
#   - demo.pdf
# quarto render index.qmd --to pdf
# quarto render index.qmd --to revealjs
---

## Key revenue forecasting questions

-   What will they owe?
-   When will they pay it?

Often it's important to consider these questions separately.


```{r}
#| label: setup

rm(list = ls())
GITROOT <- rprojroot::find_root(rprojroot::has_file(".git/index"))
source(fs::path(GITROOT, "setup.R"))

```

```{r}

pop <- readRDS(fs::path(DDATA, "ny_state_population_estimates_1970_2024.rds"))
nyspi <- readRDS(fs::path(DDATA, "bea_nyspi.rds"))
nysagi <- readRDS(fs::path(DPITNY, "nypitds2.rds"))

```


## Major challenges

-  A cloudy rear-view mirror -- hard to know the future when we don't know the present
-  Forecasters stand on shifting sands -- economic data sometimes are revised substantially
-  The economy we measure is not (quite) the economy we tax

# Where we are now

<!-- ## New York State Post-Census Population Estimates -->

```{r}
#| eval: false

p1 <- pop |>
  filter(variable == "popestimate") |>
  ggplot(aes(year, value)) +
  geom_line(colour = "blue") +
  scale_y_continuous(
    name = "millions",
    labels = scales::label_comma(scale = 1e-6, accuracy = .1)
  ) +
  scale_x_continuous(name = NULL) +
  ggtitle("Population") +
  labs(caption = "Source: U.S. Census Bureau Population Estimates Program") +
  theme_presentation_small()


```


```{r}
#| eval: false

f <- function(component, title, colour = "blue") {
  p <- pop |>
    filter(variable == component) |>
    ggplot(aes(x = factor(year), y = value)) +
    geom_col_pattern(
      fill = colour,
      position = "dodge",
      pattern = "crosshatch",
      pattern_colour = colour
    ) +
    geom_hline(yintercept = 0) +
    scale_x_discrete(name = NULL) +
    scale_y_continuous(
      name = "population change",
      labels = scales::label_comma(scale = 1, accuracy = 1),
      limits = c(-350e3, 350e3)
    ) +
    ggtitle(title) +
    labs(caption = "Source: U.S. Census Bureau Population Estimates Program") +
    theme_presentation_small()
  p
}

p2 <- f(
  "naturalchg",
  title = "Natural Change (Births minus Deaths)",
  colour = "darkgreen"
)

p3 <- f(
  "domesticmig",
  title = "Domestic Migration",
  colour = "red"
)
p4 <- f(
  "internationalmig",
  title = "International Migration",
  colour = "blue"
)

```


```{r}
#| label: four-plots
#| fig-width: 18
#| fig-height: 10
#| eval: false

(p1 + p2) / (p3 + p4)
```

## Post-Census Population Change

```{r}
#| label: pop-change-facets
#| fig-width: 18
#| fig-height: 10

keep <- c(
  "npopchg",
  "naturalchg",
  "domesticmig",
  "internationalmig"
)
labels <- c(
  "Net Population Change",
  "Natural Change (Births minus Deaths)",
  "Domestic Migration",
  "International Migration"
)

pop |>
  filter(variable %in% keep) |>
  filter(year > 2020) |>
  mutate(variable = factor(variable, levels = keep, labels = labels)) |>
  arrange(variable) |>
  ggplot(aes(x = factor(year), y = value, fill = variable)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(
      "Net Population Change" = "steelblue",
      "Natural Change (Births minus Deaths)" = "seagreen",
      "Domestic Migration" = "coral",
      "International Migration" = "mediumpurple"
    )
  ) +
  geom_hline(yintercept = 0) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(
    name = "change (thousands)",
    labels = scales::label_comma(scale = 1e-3, accuracy = 1),
    limits = c(-350e3, 350e3),
    breaks = seq(-400e3, 400e3, 100e3)
  ) +
  ggtitle("Annual Changes in New York State Population Since 2020 Census") +
  labs(caption = "Source: U.S. Census Bureau Population Estimates Program") +
  facet_wrap(~variable, ncol = 2, scales = "fixed") +
  theme_presentation() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = rel(1.3), face = "bold")
  )

```


```{r}
#| eval: false

# keep <- c(
#   "popestimate",
#   "naturalchg",
#   "domesticmig",
#   "internationalmig",
#   "npopchg",
#   "residual"
# )

# keep <- c(
#   "naturalchg",
#   "domesticmig",
#   "internationalmig"
# )

# pop |>
#   filter(variable %in% keep) |>
#   ggplot(aes(x = factor(year), y = value, fill = "blue")) +
#   geom_col(position = "dodge") +
#   geom_hline(yintercept = 0) +
#   labs(
#     title = "New York State Population Change Components",
#     x = "Year",
#     y = "Population Change",
#     fill = "Component"
#   ) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# pop |>
#   filter(variable %in% keep) |>
#   ggplot(aes(x = variable, y = value, fill = factor(year))) +
#   geom_col(position = "dodge") +
#   labs(
#     title = "New York State Population Change Components",
#     x = "Year",
#     y = "Population Change",
#     fill = "Component"
#   ) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## PIT and personal income

```{r}

nypitds2 <- readRDS(fs::path(DPITNY, "nypitds2.rds"))

```


## Employment

## Revenue

# Another section

## The economy we tax

Table or bar chart -- AGI and personal income

## Risks

- how big? / how wrong?
- when will we know

## Why capital gains are important

- large
- concentrated
- hot money - discretionary
- payment timing

## Large

## Concentrated

## Hot money

## Timing
